---
title: "NASASPACEAPP"
format: html
editor: visual
---

```{r setup, echo=FALSE}
library(ggplot2)
library(dplyr)
library(tidyr)
library(lubridate)
library(geosphere)
library(markovchain)
library(reshape2)
library(sf)
library(momentuHMM)
library(padr)
library(zoo)
library(leaflet)
library(leafem)
```

```{r data}
Full_Shark_Data <- read.csv("~/Documents/GitHub/NASASpaceApps2025/shark_data_fully_imputed_complete.csv")


```

```{r analysis}

max_gap_h <- 24  
all_states <- c("resting", "foraging", "migrating")
Shark_states_daily <- Full_Shark_Data %>%
  filter(latitude >= -90 & latitude <= 90 & longitude >= -180 & longitude <= 180) %>%
  mutate(date = as.Date(datetime)) %>%
  group_by(id, date) %>%
  summarise(
    latitude = mean(latitude),
    longitude = mean(longitude),
    depth = mean(bathymetry, na.rm = TRUE),
    chlorophyll = mean(chlorophyll, na.rm = TRUE),
    sst = mean(sst, na.rm = TRUE)
  ) %>%
  arrange(id, date) %>%
  group_by(id) %>%
  mutate(
    next_lat = lead(latitude),
    next_lon = lead(longitude),
    next_date = lead(date),
    time_diff_days = as.numeric(difftime(next_date, date, units = "days")),
    dist_m = distHaversine(cbind(longitude, latitude), cbind(next_lon, next_lat)),
    speed_mpd = dist_m / time_diff_days,  # meters per day
    state = case_when(
      !is.na(speed_mpd) & speed_mpd < 500 ~ "resting",
      !is.na(speed_mpd) & speed_mpd < 5000 ~ "foraging",
      !is.na(speed_mpd) ~ "migrating",
      TRUE ~ NA_character_
    ),
    next_state = lead(state)
  ) %>%
  filter(!is.na(state) & !is.na(next_state)) %>%
  ungroup()

complete_transitions <- expand.grid(state = all_states, next_state = all_states)

transition_counts <- Shark_states_daily %>%
  count(state, next_state) %>%
  right_join(complete_transitions, by = c("state", "next_state")) %>%
  mutate(n = ifelse(is.na(n), 0, n))

transition_prob <- transition_counts %>%
  group_by(state) %>%
  mutate(prob = n / sum(n)) %>%
  mutate(prob = ifelse(is.nan(prob), 0, prob)) %>%
  ungroup() %>%
  select(state, next_state, prob) %>%
  pivot_wider(names_from = next_state, values_from = prob)

mat <- as.matrix(transition_prob[, -1])
rownames(mat) <- transition_prob$state

row_sums <- rowSums(mat)
if (any(row_sums == 0)) {
  for (i in which(row_sums == 0)) {
    mat[i, i] <- 1
  }
}

mc <- new("markovchain", states = all_states, transitionMatrix = mat)
print("Transition Matrix:")
print(mc)

print("Example Simulation (10 steps from 'resting'):")
print(rmarkovchain(n = 10, object = mc, t0 = "resting"))

ggplot(Shark_states, aes(x = factor(state, levels = all_states))) +
  geom_bar(fill = "steelblue") +
  theme_minimal() +
  labs(title = "Frequency of Observed Shark States", x = "State", y = "Count")


mat_melt <- melt(mat)
ggplot(mat_melt, aes(Var1, Var2, fill = value)) +
  geom_tile(color = "black") +
  geom_text(aes(label = round(value, 2)), size = 5) +
  scale_fill_gradient(low = "white", high = "dodgerblue", name = "Probability") +
  labs(title = "Shark State Transition Probabilities", x = "Current State", y = "Next State") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
  
  
```
```{r second}

print("--- Step 3: Gridding the ocean and calculating rewards ---")

# Convert your daily shark data to a spatial object (sf)
# We use the full daily dataset before transitions were calculated
shark_sf_daily <- st_as_sf(Shark_states_daily, coords = c("longitude", "latitude"), crs = 4326)

# Create a spatial grid that covers the entire area of shark movements.
# You can adjust 'cellsize' to make the grid finer or coarser.
grid_cell_size <- 0.1 # in degrees
ocean_grid <- st_make_grid(shark_sf_daily, cellsize = c(grid_cell_size, grid_cell_size)) %>%
  st_as_sf() %>%
  mutate(grid_id = 1:n()) # Assign a unique ID to each grid cell

# Determine which grid cell each daily shark position falls into.
shark_grid_assignment <- st_join(shark_sf_daily, ocean_grid, join = st_within)

# Calculate the average environmental conditions for each grid cell
# that a shark has visited.
Ocean_grid_rewards <- shark_grid_assignment %>%
  as_tibble() %>% # Convert from sf for easier summarizing
  group_by(grid_id) %>%
  summarise(
    mean_chloro = mean(chlorophyll, na.rm = TRUE),
    mean_SST = mean(sst, na.rm = TRUE),
    mean_depth = mean(depth, na.rm = TRUE)
  ) %>%
  filter(!is.na(mean_chloro) & !is.na(mean_SST)) %>%
  
 
  mutate(
    chloro_norm = (mean_chloro - min(mean_chloro)) / (max(mean_chloro) - min(mean_chloro)),
    sst_norm = 1 - abs(mean_SST - 16) / (max(mean_SST) - 16),

    reward = (0.6 * chloro_norm) + (0.4 * sst_norm)
  ) %>%
  right_join(ocean_grid, by = "grid_id") %>%
  st_as_sf() %>%
  filter(!is.na(reward))


print("--- Step 4: Building the reward-based movement model ---")

grid_neighbors <- st_touches(Ocean_grid_rewards, Ocean_grid_rewards)

transition_model <- list()
for (i in 1:nrow(Ocean_grid_rewards)) {
  current_cell_id <- Ocean_grid_rewards$grid_id[i]
  neighbor_indices <- grid_neighbors[[i]]
  
  # Proceed only if the cell has neighbors
  if (length(neighbor_indices) > 0) {
    neighbor_rewards <- Ocean_grid_rewards$reward[neighbor_indices]
    neighbor_ids <- Ocean_grid_rewards$grid_id[neighbor_indices]
    
    # Probability is proportional to the reward score
    total_reward <- sum(neighbor_rewards, na.rm = TRUE)
    if (total_reward > 0) {
      probs <- neighbor_rewards / total_reward
      transition_model[[current_cell_id]] <- tibble(
        from_grid = current_cell_id,
        to_grid = neighbor_ids,
        prob = probs
      )
    }
  }
}
# Combine the list into a single, efficient data frame
transition_model_df <- bind_rows(transition_model)


print("--- Step 5: Defining the prediction function ---")

predict_next_location <- function(start_lat, start_lon, model_df, grid_sf) {

  start_point <- st_as_sf(tibble(lon = start_lon, lat = start_lat), coords = c("lon", "lat"), crs = 4326)
  
  start_cell_index <- st_nearest_feature(start_point, grid_sf)
  start_cell_id <- grid_sf$grid_id[start_cell_index]
  
  possible_moves <- model_df %>% filter(from_grid == start_cell_id)
  
  if (nrow(possible_moves) == 0) {
    print("No valid moves from this location.")
    return(NULL) 
  }
  predicted_cell_id <- sample(possible_moves$to_grid, size = 1, prob = possible_moves$prob)
  
  predicted_cell_geom <- grid_sf %>% filter(grid_id == predicted_cell_id)
  
  return(predicted_cell_geom)
}


# ===================================================================
# --- STEP 6: VISUALIZATION ---
# ===================================================================
# Goal: Create a map that integrates all our results: the reward grid,
#       the historical shark tracks, and a simulated prediction.
# -------------------------------------------------------------------

print("--- Step 6: Generating the final visualization ---")

# --- Simulate a User Click ---
# You can change these coordinates to test different starting points
user_start_lon <- -120.5
user_start_lat <- 34.5
user_start_point <- st_as_sf(tibble(lon = user_start_lon, lat = user_start_lat), coords = c("lon", "lat"), crs = 4326)

# --- Make a Prediction using the function from Step 5 ---
predicted_cell <- predict_next_location(user_start_lon, user_start_lat, transition_model_df, Ocean_grid_rewards)

# Proceed to plot only if a prediction was successfully made
if (!is.null(predicted_cell)) {
  predicted_centroid <- st_centroid(predicted_cell)

  # --- Create the Plot using ggplot ---
  final_map <- ggplot() +
    # Layer 1: The reward grid (background)
    geom_sf(data = Ocean_grid_rewards, aes(fill = reward), color = NA, alpha = 0.7) +
    scale_fill_viridis_c(name = "Cell Reward Score") +
    
    # Layer 2: The historical daily shark positions, colored by behavior
    geom_sf(data = shark_sf_daily, aes(color = state), size = 2, alpha = 0.8) +
    scale_color_manual(name = "Behavior (Rule-Based)", values = c("resting" = "orange", "foraging" = "green", "migrating" = "dodgerblue")) +
    
    # Layer 3: The user's click and the model's prediction
    geom_sf(data = user_start_point, color = "red", shape = 4, size = 5, stroke = 1.5) +
    geom_sf(data = predicted_centroid, color = "white", shape = 3, size = 5, stroke = 1.5) +
    
    # Layer 4: An arrow connecting the click to the prediction
    geom_segment(aes(x = user_start_lon, y = user_start_lat,
                     xend = st_coordinates(predicted_centroid)[1], yend = st_coordinates(predicted_centroid)[2]),
                 arrow = arrow(length = unit(0.3, "cm")), color = "black", size = 1.2) +
    
    # Labels and theme
    labs(title = "Predictive Shark Movement Model",
         subtitle = "Click (Red X) -> Predicted Next Day Location (White +)",
         x = "Longitude", y = "Latitude") +
    theme_minimal()

  print(final_map)
}

print("--- Workflow Complete ---")
```

```{r shiny_app}
# ===================================================================
# --- STEP 7: INTERACTIVE SHINY APPLICATION ---
# ===================================================================
# Goal: Take all the processed data and models and build an
#       interactive map where the user can click to get a prediction.

# --- SHINY UI (USER INTERFACE) ---
# Defines the layout and appearance of the web app.
ui <- fluidPage(
  titlePanel("Interactive Shark Movement Prediction"),
  
  sidebarLayout(
    sidebarPanel(
      h4("Instructions"),
      p("Click anywhere on the map to select a starting location for a shark."),
      p("The model will predict the most likely grid cell the shark will move to on the next day, based on historical movements and environmental rewards (chlorophyll, SST)."),
      hr(),
      h4("Legend"),
      p(HTML("🟠 <b>Resting:</b> Low-speed movement.")),
      p(HTML("🟢 <b>Foraging:</b> Medium-speed movement.")),
      p(HTML("🔵 <b>Migrating:</b> High-speed, long-distance movement.")),
      p(HTML("❌ <b>Your Click:</b> The starting point for the prediction.")),
      p(HTML("➕ <b>Prediction:</b> The predicted location for the next day."))
    ),
    mainPanel(
      # The leaflet map will be rendered here and will fill the main panel
      leafletOutput("map", height = "85vh")
    )
  )
)

server <- function(input, output, session) {


  output$map <- renderLeaflet({
    leaflet() %>%
      addProviderTiles(providers$Esri.OceanBasemap, group = "Ocean Basemap") %>%
      addProviderTiles(providers$CartoDB.Positron, group = "Light Basemap") %>%
      

      addPolygons(
        data = Ocean_grid_rewards,
        fillColor = ~colorNumeric("viridis", reward)(reward),
        fillOpacity = 0.6,
        stroke = FALSE,
        group = "Reward Grid",
        popup = ~paste("Reward Score:", round(reward, 2))
      ) %>%
      

      addCircleMarkers(
        data = shark_sf_daily,
        color = ~case_when(
          state == "resting" ~ "orange",
          state == "foraging" ~ "green",
          state == "migrating" ~ "dodgerblue",
          TRUE ~ "grey"
        ),
        radius = 4,
        stroke = TRUE, weight = 1, color = "black",
        fillOpacity = 0.8,
        group = "Historical Tracks",
        popup = ~paste("Date:", date, "<br>Behavior:", state)
      ) %>%
      

      addLayersControl(
        baseGroups = c("Ocean Basemap", "Light Basemap"),
        overlayGroups = c("Reward Grid", "Historical Tracks"),
        options = layersControlOptions(collapsed = FALSE)
      ) %>%
      addScaleBar()
  })
  observeEvent(input$map_click, {
    click <- input$map_click
    lat <- click$lat
    lon <- click$lng

    predicted_cell <- predict_next_location(lat, lon, transition_model_df, Ocean_grid_rewards)
    

    proxy <- leafletProxy("map")
    
    # First, clear any markers and arrows from the previous prediction
    proxy %>% clearGroup("prediction_group")
    
    # Add a marker for the user's click
    proxy %>% addAwesomeMarkers(
      lng = lon, lat = lat,
      icon = awesomeIcons(icon = "times", library = "fa", markerColor = "red"),
      group = "prediction_group"
    )
    
    if (!is.null(predicted_cell)) {
      predicted_centroid <- st_centroid(predicted_cell)
      pred_coords <- st_coordinates(predicted_centroid)
      
      proxy %>% addAwesomeMarkers(
          lng = pred_coords, lat = pred_coords,
          icon = awesomeIcons(icon = "plus", library = "fa", markerColor = "purple"),
          group = "prediction_group"
        ) %>%
        addArrowhead(
          lng0 = lon, lat0 = lat,
          lng1 = pred_coords, lat1 = pred_coords,
          color = "black", weight = 2, opacity = 1,
          group = "prediction_group"
        )
    }
  })
}


shinyApp(ui, server)